High-Performance Language Primitives for Distributed Architectures using Asynchronous Events

### 2012-07-22 doc/asplos2013/impl.tex
first draft of implementation, playing around a bit in interface too

@@

>>>>

Like events, locks are created on demand by the node on which the creation request was made.  The lock
handles are also statically divided across the nodes, allowing the creation to be done without inter-node
communication.  Unlike events whose ownership is static, locks may migrate, so although the {\em creating}
node is the initial owner of a lock, that ownership can be transferred to other nodes.

Locks are often used to mediate access to data, and the size of that data is often relatively small.  To
avoid adding the latency of inter-node data copies to critical sections, a lock may also be created with a
{\em payload}, a small (i.e. less than 4KB - the limit for an active message's payload) piece of data that is moved around with the lock and guaranteed to be coherent if
the lock is held.

Since any node may at some point be the owner of a lock, all nodes use the same data structure to track the
state of the lock (they differ only in when the structure is allocated -
nodes other than the creator still wait until the first reference to the lock on that node).  The structure
tracks the following:
\begin{itemize}
\item {\em owner node} - the most recently known owner of the lock.  If the current node is the owner, this
information is always accurate.  If not, this information may be stale, but at least the recorded ``owner''
is sure to have more up-to-date information.
\item {\em lock status} - records whether the lock is currently held.  Only the current lock owner has
guaranteed-current information here.
\item {\em local waiters} - a list of pending local lock requests.  This data is always valid as well.
\item {\em remote waiters} - a bitmask of which other nodes have pending requests.  This bitmask is only
used when the current node is the owner of the lock.
\item {\em local payload pointer and size} - the local node's copy of the lock payload
\end{itemize}

When a lock request is made, an event is created to track when the grant occurs.  The current node then
examines its copy of the lock data structure (creating it if
this the first operation for that lock on that node).  If the current node is the owner and the lock isn't
held, the lock is granted immediately and the event is triggered.  If the lock is held, the event is added
to the list of local waiters.  (This is the only thing that is stored.  The actual ``requestor'' of the lock
is implicitly captured in which operations are dependent on the lock grant event.)  If the current node
isn't the owner, a {\em lock request} active message is
sent to the most recently known owner.  (If that information proves to be stale, the request is forwarded
by the receiving node, eventually catching up to the current owner.)  If the current owner's status shows
the lock is currently held, the requesting node is added to the remote waiters bitmask.  If the lock is
not held, the ownership of the lock is given to the requesting node via a {\em lock transfer} active
message.  A lock transfer message includes the current remote waiters bitmask (in case other nodes are
have also requested the lock) and an up-to-date copy of the lock's payload.

Similarly, an unlock request is first checked against the local node's lock state.  If the local node is
not the owner, an {\em unlock} active message is sent to the most recently known owner, which is forwarded
if necessary.  Once the unlock request is on the lock's current owning node, the local waiter list is
examined.  If the list is non-empty, the lock remains in the ``locked'' state and the first lock grant
event is pulled off the local waiter list and triggered.  If instead the local waiter list is empty, the
lock state is changed to ``unlocked'', and the remote waiter bitmask is examined.  If any bits are set, then
one of them is chosen and the corresponding node becomes the new owner via a lock transfer active message.

The unfairness that results from a lock favoring local waiters over remote waiters is intentional.  When the
contention on a lock is high (the only time when the question of fairness is even relevant), the latencies
involved in transferring a lock between nodes can be the limiter on throughput.  By minimizing the number
of lock transfers, the throughput on the lock is maximized.

A similar issue can arise with the forwarding of lock request active messages.  If two nodes are
transferring a lock very quickly back and forth, a third node's request could be forwarded an arbitrary
number of times as it chases the lock around.  Although this can result in unfairness, it doesn't 
prevent forward progress.  A lock is only transferred to a node with at least one local waiter, so if 
the third node's request ever became the only one left, the lock transferring would stop and the request
would finally catch up.

@@

In addition to normal physical regions, our implementation supports two different kinds of reduction-only
physical regions, designed to optimize the common pattern in which multiple tasks make commutative 
{\em reductions} into shared data.  Instead of having to move the existing data to a suitable memory and
then applying individual reduction operations to the existing data 
right away (as would happen in a normal physical region), reduction instances allow a large number of
reduction operations to be batched up and transferred efficiently to the original location of the data.
This decoupling from the original location of the data can allow allow greater parallelism
in cases where there are available processors that cannot all access the same memory.

The first kind of reduction-only physical region is a {\em reduction fold instance}, which may only be
used for reduction operations that support a {\em fold} operation.  A reduction fold instance is similar
to a normal physical region in that it is implemented as an array, indexed by the same element indices used
for the normal physical regions.  The difference is that each array element is a value of the reduction's
{\tt RHS} type instead of the actual element type of the region (i.e. the {\tt LHS} type).  When the client
requests a reduction operation to a particular location, the supplied right-hand-side value is folded in
to the corresponding location in the array.  An arbitrary number of reductions can be done to each location.
When the reduction fold instance is copied back to a normal instance, the contents of the reduction instance
(which are generally the same size or smaller than the normal region contents) are transferred in a large
block to the location of the normal instance, where an application of the reduction function is performed
once for each location in the region via a cache-friendly linear sweep over the memory.

The second kind of reduction instance is a {\em reduction list instance}.  Although it generally doesn't
perform as well as a reduction fold instance, it can be better for cases with sparse updates.  It is also
the only option when a reduction operation doesn't support a {\em fold} operation.  Instead of folding
together multiple reductions to the same location, a reduction list instance keeps a log of every
reduction operation requested (i.e. the location and the right-hand-side value).  When the reduction list 
instance is copied back to a normal physical region, the whole list is transferred and then replayed at
the target region's location.  Except in sparse update cases, this list is often larger than the target
region, and any randomness in the sequence of locations in the list can impact the memory access efficiency
of the accesses to the target region.  However, as we will show in section~\ref{sec:microbenchmarks}, 
the benefits of enabling parallelism and hiding latency usually more than make up for these drawbacks.

### 2012-07-22 doc/asplos2013/interface.tex
first draft of implementation, playing around a bit in interface too

\begin{lstlisting}[float={t},label={lst:lockapi},caption={Deferred Lock Interface and Example.}]
class Lock {
public:
  const unsigned id;
  Event lock(unsigned mode = 0, bool exclusive, 
              Event wait_on = Event::NO_EVENT) const;
  void unlock(Event wait_on = Event::NO_EVENT) const;
  // Create a new lock, destroy an existing lock
  static Lock create_lock();
  void destroy_lock();
};

@@

Clients can also associate an operation as part of a physical region.  One example of
associating an operation with a physical region is shown on lines 13-14.  This instance
of the {\tt create\_instance} method
will create a physical region called a {\em reduction instance} that will be assoicated 
with the given {\tt ReductionOp} specified by the ID.  The interface for a {\tt ReductionOp} is shown on 
lines 26-33.  A reduction operation must have an {\tt apply} method, and can optionally 
implement two additional methods, {\tt init} and {\tt fold} that support additional 
optimizations described in Section~\ref{subsec:reducimpl}.  The two template parameters 
describe the base element type of the region ({\tt LHS}:left-hand-side) and the type of 
the argument to the reduction ({\tt RHS}:right-hand-side).  The reduction operation has 
total control over the choice of data layout for the reduction instance (not shown).
Reduction instances can be copied to other reduction instances with the same operation
or to regular physical regions of the same class.

@@

template<typename LHS, typename RHS>
class ReductionOp {
public:
  void apply(LHS *lhs, RHS rhs);
  // Optional methods for optimizations
  void init(RHS *rhs, size_t num_elmts);
  void fold(RHS *rhs1, RHS rhs);
};

>>>>

\begin{lstlisting}[float={t},label={lst:lockapi},caption={Deferred Lock Interface and Example.}]
class Lock {
public:
  const unsigned id;
  Event lock(Event wait_on = Event::NO_EVENT) const;
  void unlock(Event wait_on = Event::NO_EVENT) const;
  // Create a new lock, destroy an existing lock
  static Lock create_lock(size_t payload_size = 0);
  void *payload_ptr();
  void destroy_lock();
};

@@

If the only operations that will be performed on locations in a {\tt RegionInstance} object,
a special reduction version can be created by providing the ID of a {\em reduction operator}
when the object is created (line 13).  Reduction-only {\tt RegionInstance} objects can 
increase the available parallelism and hide latency better, as we will show in section~\ref{sec:microbenchmarks}.  The interface supports any reduction operation that
is commutative (i.e. multiple operations to the same location can be reordered without 
changing the final result) and is coded as an {\tt apply} method as shown on line 29.
Many commonly-used reduction operations are also {\em foldable} (e.g. $(l \text{ += } r_1) \text{ += } r_2$ can
be folded as $l \text{ += } r_1 + r_2$), and if the {\tt fold} operation and a {\em right identity}
(e.g. $0$ is a right identity for subtraction because $x - 0 = x$ for all $x$) are provided,
the runtime can use a {\em reduction fold instance}, described in section~\ref{subsec:reducimpl},
for ever better performance.  Although the way they store data is very different,
reduction-only instances can be copied to other reduction-only instances (or normal 
instances) of the same class with the same asynchronous copy operations.

%% Clients can also associate an operation as part of a physical region.  One example of
%% associating an operation with a physical region is shown on lines 13-14.  This instance
%% of the {\tt create\_instance} method
%% will create a physical region called a {\em reduction instance} that will be assoicated 
%% with the given {\tt ReductionOp} specified by the ID.  The interface for a {\tt ReductionOp} is shown on 
%% lines 26-33.  A reduction operation must have an {\tt apply} method, and can optionally 
%% implement two additional methods, {\tt init} and {\tt fold} that support additional 
%% optimizations described in Section~\ref{subsec:reducimpl}.  The two template parameters 
%% describe the base element type of the region ({\tt LHS}:left-hand-side) and the type of 
%% the argument to the reduction ({\tt RHS}:right-hand-side).  The reduction operation has 
%% total control over the choice of data layout for the reduction instance (not shown).
%% Reduction instances can be copied to other reduction instances with the same operation
%% or to regular physical regions of the same class.

@@

template<typename LHS, typename RHS>
class ReductionOp {
public:
  static void apply(LHS *lhs, RHS rhs);
  // Optional methods for folding optimization
  static void fold(RHS *rhs1, RHS rhs2);
  static const RHS right_identity;
};
